Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-07 tex/main.tex
Minor fixes

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcalx{M}}
\newcommand{\I}{\mathcalx{J}} %%% there's no difference, right? --JB
%% (I find \mathcalx{I} ugly, with its two horizontal bars, hence the J)
\newcommand{\J}{\mathcalx{J}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
%%% @ANDY: I'm trying "A" like "abstraction", again in sync with the abstract--concrete
%%% terminology from abstract interpretation
\newcommand{\conv}{\mathcalx{A}}

@@

\newcommand{\boolop}{\con{b}}
\newcommand{\forallf}[1]{\forall_{\!#1\:}}
\newcommand{\fnull}{\emptyset}
\newcommand{\vecfarg}[1]{\vec{\concret}_{#1}}
\newcommand{\farg}[1]{\concret_{#1}}
\newcommand{\fargx}[2]{\concret^{\vthinspace#2}_{#1}}
\newcommand{\fargtype}[1]{\alpha_{#1}}

@@

\begin{lemma}\label{lem:conv}
If $\varphi$ is a formula not containing function definitions,
and $\I$ is a $\extendsig{\Sigma}$-interpretation satisfying $\conv( \varphi )$,
then $\I$ satisfies a set of literals $L \cup X$ such that
$L$ is a set of $\Sigma$-literals that entail $\varphi$,
and $X$ is the set of literals
$\{ \exists a : {\fargtype{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t ) \mid \con{f}( \vec t ) \in \terms( L ), \con{f} \in \sfundefs{\Sigma} \}$.
\end{lemma}

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissible.
\ct{
You cannot possibly mean "if and only if" in the two sentences above.
Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
(1) is sat iff (2) is sat. That is immediate.
}
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargtype{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $s( c )$,
any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

The \texttt{--fmf-fun} option instructs \cvc to assume that the
functions provided by the user are admissible.
Admissibility must be checked separately \emph{by the user}---for example, using a
syntactic criterion or a termination prover.
If this criterion is not met, \cvc may answer \emph{sat} on
an unsatisfiable problem. For example, if we add the inconsistent definition
%
\begin{quote}
\small
\begin{verbatim}
(define-fun-rec e ((n Int)) Int (+ (e n) 1))
\end{verbatim}
\end{quote}
%
to the above problem, \cvc will still produce the same output
if \texttt{--fmf-fun} is specified.

>>>>

%\newcommand{\functerms}{\mathbf{DFT}}
\newcommand{\vars}{\mathbf{V}}
\newcommand{\M}{\mathcalx{M}}
\newcommand{\I}{\mathcalx{I}} %%% there's no difference, right? --JB  need to simultaneously refer to 2 different interpretations --ajr.
%% (I find \mathcalx{I} ugly, with its two horizontal bars, hence the J)
\newcommand{\J}{\mathcalx{J}}
\def\AIF{\qtab\keyword{if}\ }
\def\THEN{\ \keyword{then}\ }
\def\AELSE{\untab\qtab\keyword{else}\ }
\def\FI{\untab}
\def\RETURN{\keyword{return}\ }
\def\ENDPROC{\untab}
%%% @ANDY: I'm trying "A" like "abstraction", again in sync with the abstract--concrete
%%% terminology from abstract interpretation
\newcommand{\conv}{\mathcalx{A}}

@@

\newcommand{\boolop}{\con{b}}
\newcommand{\forallf}[1]{\forall_{\!#1\:}}
\newcommand{\fnull}{\emptyset}
\newcommand{\vecfarg}[1]{\vec{\concret}_{#1}}
\newcommand{\farg}[1]{\concret_{#1}}
\newcommand{\fargx}[2]{\concret_{#1,\vthinspace#2}}
\newcommand{\fargtype}[1]{\alpha_{#1}}

@@

\begin{lemma}\label{lem:conv}
If $\varphi$ is a formula not containing function definitions,
and $\I$ is a $\extendsig{\Sigma}$-interpretation satisfying $\conv( \varphi )$,
then $\I$ satisfies a set of $\Sigma$-literals $L$ that entail $\varphi$,
and $\I$ satisfies
$\{ \exists a : {\fargtype{\con{f}}}. ( \vecfarg{\con{f}}( a ) \teq \vec t ) \mid \con{f}( \vec t ) \in \terms( L ), \con{f} \in \sfundefs{\Sigma} \}$.
\end{lemma}

@@

\begin{example}
Let us revisit the translated formula~(\ref{eq:ex-after}) from
Example~\ref{ex:translation}. This formula is unsatisfiable if and only if
formula~(\ref{eq:ex-before}) is unsatisfiable since $\fargtype{f}$ can be
interpreted as $\typeint$, and $\farg{f}$ can be interpreted as the identity
function. Conversely, we claim formula~(\ref{eq:ex-after}) is satisfiable if
and only if formula~(\ref{eq:ex-before}) is satisfiable, noting that the set
$\{ \forallf{\con{s}} x\typ{\typeint}.\;\, \lite\bigl( x \leq 0,\allowbreak\; \con{s}(x)
\teq 0,\; \con{s}( x ) \teq x + \con{s}( x-1 ) \bigr) \}$ is admissible.
\ct{
You cannot possibly mean "if and only if" in the two sentences above.
Once I show that (1) is unsat iff (2) is unsat, I do not need to argue that
(1) is sat iff (2) is sat. That is immediate.
}
As stated
in the proof of Theorem~\ref{thm:equi}, the reason is that any model of
formula~(\ref{eq:ex-after}) must satisfy a formula that is closed under
function expansion, hence implying that formula~(\ref{eq:ex-before}) has a
model. In more detail, let $\varphi_\con{s}$ be the formula $\forallf{\con{s}} x.\;\,
\lite\bigl( x \leq 0, \con{s}(x) \teq 0, \con{s}( x ) \teq x + \con{s}( x-1 )
\bigr)$. The rightmost conjunct $\exists a : {\fargtype{\con{s}}}.
\farg{\con{s}}( a ) \teq \con{c}$ ensures that any model of
formula~(\ref{eq:ex-after}) satisfying $\con{s}( \con{c} ) > 100$ also
satisfies $\varphi_{\con{s}}[\con{c}/x]$. The nested existential
$\exists b.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1$ ensures that
a similar constraint is enforced recursively: when $\varphi_{\con{s}}[t/x]$ is used in part to evaluate $\con{s}( \con{c} )$,
any model of $\varphi_{\con{s}}[t/x]$ must also satisfy $\varphi_{\con{s}}[t-1/x]$ when $t > 0$. Given a model $\I$ of
formula~(\ref{eq:ex-after}) that interprets $\con{c}$ as $14$, it is easy to see that
due to the structure of formula~(\ref{eq:ex-after}),
$\I$ must satisfy the following formula $\psi$:
%
\begin{equation} \label{eq:fun-closure}
\begin{split}
& \con{s}( \con{c} ) > 100 \wedge %\displaystyle
  \bigwedge\nolimits_{\vthinspace i=0}^{\!13} \bigl( \neg (\con{c}-i \leq 0) \wedge \con{s}( \con{c}-i ) \teq i + \con{s}( \con{c}-i-1 ) \bigr) \\[-\jot]
{\wedge}\; & \con{c}-14 \leq 0 \wedge \con{s}( \con{c}-14 ) \teq 0
\end{split}
\end{equation}
%
This formula is closed under function expansion,
since it contains the $\con{s}$ applications corresponding to
$\con{s}( \con{c}-i )$ for $i = 0, \ldots, 14$. The formula also
entails $\varphi_\con{s}[\con{c}-i/x]$ for $i = 0, \ldots, 14$.
Since $\varphi_\con{s}$ is an admissible definition,
there must exist a $\Sigma$-interpretation satisfying $\psi \wedge \varphi_\con{s}$
which entails formula~(\ref{eq:ex-before}).\xend
\end{example}

@@

The \texttt{--fmf-fun} option instructs \cvc to assume that the
functions provided by the user are admissible.
Admissibility must be checked separately \emph{by the user}---for example, using a
syntactic criterion or a termination prover.
If this criterion is not met, \cvc may answer \emph{sat} on
an unsatisfiable problem. For example, if we add the inconsistent definition
%
\begin{quote}
\small
\begin{verbatim}
(define-fun-rec e ((n Int)) Int (+ (e n) 1))
\end{verbatim}
\end{quote}
%
to the above problem, \cvc will still return ``satisfiable"
if \texttt{--fmf-fun} is specified.
